Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2024-12-26 02-antecedentes.qmd
+ ajustes antecedentes

The justification of economic inequalities is a controversial field of study. As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable. Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as “the defensibility of unequal relations between people” [-@barry_theories_1989, p. 3], whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification [@trump_income_2018;@kelley_legitimate_2021;@mijs_paradox_2021]. Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.

@@

Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023]. For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps. However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.

[[ver si esto queda acá o baja]] The perception of economic inequality is one of the factors associated with the justification of educational inequality. As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023]. For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S. points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality. Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education. Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]

@@

As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification within this time frame. Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021]. Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been affected in this period. Consistently, our first hypothesis is:

@@

Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019], expecting that:

@@

Thirdly, we address inequality perception. As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized. As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:

@@

Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time. As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged. As a consequence, the justification of inequality could show a decrease over time:

>>>>

The justification of economic inequalities is a controversial field of study. As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable. Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as “the defensibility of unequal relations between people” [-@barry_theories_1989, p. 3], whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification [@trump_income_2018;@kelley_legitimate_2021;@mijs_paradox_2021]. Both theoretical and empirical approaches have been able to establish a clear distinction between inequality and justice, opening the question of to what extent inequality can be tolerated or even justified.

@@

@@

As there are measures before and after the pandemic period, firstly we explore to what extent it is possible to identify variations in inequality justification within this time frame. Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021]. Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been affected in this period. Consistently, our first hypothesis is:

@@

Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019]. Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023]. For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps. However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification. In line with this evidence, we expect that:

@@

Thirdly, we address inequality perception. As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023]. For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S. points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality. Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education. Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]. As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:

@@

Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time. As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged in the time frame analyzed. As a consequence, the justification of inequality could show a decrease over time:
